Some Graphical Realizations of Two-Row Specht Modules of Iwahori-Hecke Algebras of the Symmetric Group

### 2019-07-28 tex/Rough-Overview.tex

\documentclass{amsart}   
\usepackage{RepStyle} 
\usepackage{NatMacros} %********REPLACE WITH MilesMacros TO KILL RELIANCE ON \ifnum AND HAVE INCOMPLETE OUTPUT***********bleh
\begin{document}

\title[Some Graphical Realizations of Two-Row Specht Modules of Hecke Algebras]{Some Graphical Realizations of Two-Row Specht Modules  of Hecke Algebras}
\author[Miles Johnson \& Natalie Stewart]{Miles Johnson \& Natalie Stewart\\
  Mentor Oron Propp\\
Project Suggested by Roman Bezrukavnikov\\ \; \\
July 31, 2019
}

\begin{titlepage}
  \maketitle
  \begin{abstract}
  Abstract goes here
  \end{abstract}
\end{titlepage}

@@

\subsection{Crossingless Matchings}
\begin{definition}
  A \emph{crossingless matching on $2n+r$ indices with $r$ anchors} is a partition of $\cbr{1,\dots,2n+r}$ into $n$ parts of size $2$ and $r$ of size 1 such that no two parts of size two``cross'', i.e. there are no parts $(a,a')$ and $(b,b')$ such that $a < b < a' < b'$, and no parts of size one are ``inside'' of a part of size two, i.e. there are no $c, (a,a')$ such that $a < c < a'$.
  We will call these arcs and anchors, respectively.
  Then, define $W^r_{2n+r}$ to be the $k$-vector space with basis the set of generalized crossingless matchings on $2n+r$ indices with $r$ anchors.

@@

We will prove that $W := W_{2n+r}^r$ and $S := S^{(n+r,n)'}$ are isomorphic as representations in the case that $e > n + r + 1$.
Note that, when $r = 0$, these have the same dimension given by the $n$th catalan number $C_n$.

\subsection{Fibonacci Representation}
\def\vara{\alpha}
\def\varb{\beta}
\def\varc{\gamma}
\def\vard{\delta}
\def\vare{\varepsilon}
\def\vs{\texttt{*}}
\def\vp{\texttt{0}}
Now suppose that $e = 5$ and $k$ contains the algebraic number $(-1 - q^2 - q^3)^{3/2}$.
Let $V^m$ be a $k$-vector space with basis given by the strings $\cbr{\vs,\vp}^{n+1}$ such that the character $\vs$ never appears twice in a row. 
We will suppress the superscript whenever it is clear from context.

@@

\newpage
\section{Preliminaries on Specht Modules}
\begin{proposition}
  \begin{enumerate}[label={(\roman*)}]
    \item
    Suppose that $n,r > 0$.
    Then, a filtration of $\Res W_{2n + r}^r$ is given by
    \begin{equation}
      0 \subset W_{2n + r - 1}^{r-1} \subset \Res \, W_{2n + r}^r \label{Filtration}
     \end{equation}
     with $\Res \, W_{2n + r}^r / W_{2n + r - 1}^{r-1} \simeq W_{2n + r - 1}^{r + 1}$. 
    \item
      We have the following isomorphism of representations:
      \begin{equation}
        W_{2n - 1}^1 \simeq \Res \; W_{2n}^0 \label{0 Restriction}
       \end{equation}
  \end{enumerate}
\end{proposition}
\begin{proof}
  \textbf{(i)}
  Note that we may identify the subrepresentation of $\Res \; W_{2n + r}^r$ having anchor $n$ with $W_{2n + r - 1}^{r-1}$.

\begin{lemma}
  Suppose $n,r > 0$.
  Then, we may characterize the socle of $\Res \, D^\lambda$ as follows:
  \[
    \soc\prn{\Res \, D^{(n+r,n)'}} \simeq \begin{cases}
      D^{(n + r - 1,n)'} & e \mid r+2\\
      D^{(n + r,n-1)'} & e \nmid r+2, \; e \mid r\\
      D^{(n+r-1,n)'} \oplus D^{(n+r,n-1)} & e \nmid r+2,r
    \end{cases}
  \]
\end{lemma}
\begin{proof}
  This amounts to computations of the hook lengths $\beta(1,2)$ and $\gamma(1,2)$:
  \begin{align*}
    \beta_\lambda(1,2) &= r + 2\\
    \gamma_\lambda(1,2) &= r
  \end{align*}
  Since $2$ is the largest removable number, $D^{(n+r,n-1)'} \subset D^{(n+r,n)'}$ iff $e \nmid r + 2$.
  Further, if $e \nmid r + 2$, then $D^{(n+r-1,n)'} \subset D^{(n+r,n)'}$ iff 1 is good iff $e \nmid r$
\end{proof}

\begin{lemma}\label{S irreducibility}
  \begin{enumerate}[label={(\roman*)}]
    Assume $e > 2$ and $n > 0$.
    \item Suppose $e > n$.
      Then, $S^{(n+r,n)'}$ is irreducible iff $e \nmid l$ for all $r + 2 \leq l \leq n + r + 1$.
    \item Suppose $e \leq n$.
      Then, $S^{(n+r,n)'}$ is irreducible iff $e | r + 1$ and $\nu_{e,p}(h_{ab}^\lambda)$ acquires exactly two values over rows $a \leq n$.
  \end{enumerate}  
\end{lemma}
\begin{proof}
  \textbf{(i)}
  Note that $\nu_p(l) \neq -1$ for all $l$ and only hook lengths in the first column may vanish mod $e$;
  hence we may equivalently prove that $e$ divides no hook lengths in the first $n$ rows of the first column.
  These hook lengths are precisely $r + 2,\dots,n+r+1$.

@@

@@

@@

\section{Crossingless Matchings and Specht Modules}
\subsection{Preliminaries on Crossingless Matchings}
\begin{lemma}
  \label{Cyclic}
  Every basis vector in $W_{2n + r}^r$ is cyclic.
\end{lemma}
\begin{proof}
  We have already proven this in the $r = 0$ case, so suppose that $r > 0$.

@@

Let $K := \bigcap_{i=1}^{2n+r-1} \ker (1 + T_i) = \ker \bigoplus_{i = 1}^{2n + r - 1} (1 + T_i)$.
This will be a large technical tool in our proof of irreducibility.
\begin{lemma}\label{Represented}
  Let $w_j$ be the basis vector with anchors $1,\dots,r$ and all arcs of maximal length.
  Suppose $w \in K \backslash \cbr 0$.
  Then, $w_j$ is represented in $w$.
\end{lemma}
\begin{proof}
  We will show this in steps; first, we show that, given that a vector is represented with anchors $1,\dots,s$, there must be a vector represented in $w$ with $(s+1)$st anchor, including when $s = 0$;
  this implies that a vector is represented with anchors $1,\dots,r$.
  Then, we will show that, given a vector is represented with anchors $1,\dots,r$ and first $s$ arc-lengths $n,n-2,\dots,n-2s$, there is a vector represented with these and the $(s+1)$st arc-length $n-2s-2$.
  This implies that $w_j$ is represented.

@@

  \vspace{\lemsep}
  \textit{Step 2.}
  This step is similar;
  suppose that $s < n$ is the maximal number such that a vector with anchors $1,\dots,r$ and first $s$ arc-lengths $n,\dots,n-2s$ is represented.
  Take the vector $w_i$ which, among vectors represented in $w$ with anchors $1,\dots,r$ and first $s$ arc-lenths $n,\dots,n-2s$, has maximal length $t$ of the arc beginning at index $r + s + 1$.
  Then, $q^{1/2}(1 + T_{r + s + t})w_i$ is mapped to only by $w_i$ and vectors having anchors $1,\dots,r$ and first $s+1$ arc-lengths $n,\dots,n-2s,t'$ with $t' > t$, which are not represented in $w$;
  hence $q^{-1/2}(1 + T_{r + s + t})w_i$ is represented in $(1 + T_{r+s+t})w$, giving $w \notin \ker (1 + T_{r + s + t})$.
  The $s = 0$ case is similar, and implies that $s = n$.
\end{proof}

\begin{lemma}\label{Trivial}
  Suppose $e \nmid n + r + 1$.
  Then, $K = 0$.
\end{lemma}
\begin{proof}
  Consider the matrix $A = \bigoplus (1 + T_i)$ having kernel $K$.
  It is sufficient by lemma \ref{Represented} to show that $A$ includes a row $[0,\dots,0,1,0,\dots,0]$ with a nonzero entry only on the column $j$.

@@

  We may prove invertibility of this matrix by proving that $\det B_{n + r} = \brk{n + r + 1}_q$ inductively on $n+r$.
  This is satisfied for our base case $n + r = 1$, so suppose that it is true for each $m < n + r$.
  Then,
  \begin{align*}
    \det B_{n+r} &= (q + 1) \det B_{n + r - 1} - q \det B_{n + r - 2}\\
    &= (q + 1)(1 + \dots + q^{n + r - 1}) - (q + \dots + q^{n + r - 1})\\
    &= 1 + \dots + q^{n + r}\\
    &= \brk{n + r + 1}_q.
  \end{align*}
  Hence $\det B_{n + r} \neq 0$, and $K = 0$.
\end{proof}

\subsection{Correspondence}
\begin{proposition}
  The representation $W_{2n + r}^r$ is irreducible when $e > n + r + 1$. 
\end{proposition}
\begin{proof}
  We proceed by induction on $2n + r$.
  Note that, by identification with the trivial and sign representations, the base case $2n + r = 2$ is already prove, so suppose we have proven this for each $2m + s < 2n + r$.

@@

\begin{corollary}
  Suppose $n,r > 0$ and $e > n + 1$.
  Then, the sequence \eqref{Filtration} is a composition series of $\Res \, W_{2n + r}^r$.\qed
\end{corollary}

@@

@@

\section{Acknowledgements}

\begin{thebibliography}{9}

@@

\end{thebibliography}

>>>>

\documentclass{amsart}   
\usepackage{RepStyle} 
\usepackage{NatMacros}

% Emulates \subsection* but doesn't add to ToC (I don't know why amsart is weird 
\newcommand{\fakesubsection}[1]{
  \vspace{7pt}
  \noindent \textbf{#1.}
}

\makeatletter
%Table of Contents
\setcounter{tocdepth}{3}

% Add bold to \section titles in ToC and remove . after numbers
\renewcommand{\tocsection}[3]{%
  \indentlabel{\@ifnotempty{#2}{\bfseries\ignorespaces#1 #2\quad}}\bfseries#3}
% Remove . after numbers in \subsection
\renewcommand{\tocsubsection}[3]{%
  \indentlabel{\@ifnotempty{#2}{\ignorespaces#1 #2\quad}}#3}
\let\tocsubsubsection\tocsubsection% Update for \subsubsection
%...

\newcommand\@dotsep{4.5}
\def\@tocline#1#2#3#4#5#6#7{\relax
  \ifnum #1>\c@tocdepth % then omit
  \else
    \par \addpenalty\@secpenalty\addvspace{#2}%
    \begingroup \hyphenpenalty\@M
    \@ifempty{#4}{%
      \@tempdima\csname r@tocindent\number#1\endcsname\relax
    }{%
      \@tempdima#4\relax
    }%
    \parindent\z@ \leftskip#3\relax \advance\leftskip\@tempdima\relax
    \rightskip\@pnumwidth plus1em \parfillskip-\@pnumwidth
    #5\leavevmode\hskip-\@tempdima{#6}\nobreak
    \leaders\hbox{$\m@th\mkern \@dotsep mu\hbox{.}\mkern \@dotsep mu$}\hfill
    \nobreak
    \hbox to\@pnumwidth{\@tocpagenum{\ifnum#1=1\bfseries\fi#7}}\par% <-- \bfseries for \section page
    \nobreak
    \endgroup
  \fi}

\def\l@subsection{\@tocline{2}{0pt}{2.5pc}{5pc}{}}
\makeatother

\begin{document}

\title[Some Graphical Realizations of Two-Row Specht Modules of Hecke Algebras]{Some Graphical Realizations of Two-Row Specht Modules of Iwahori-Hecke Algebras of the Symmetric Group}
\author[Miles Johnson \& Natalie Stewart]{Miles Johnson \& Natalie Stewart\\
  Mentor Oron Propp\\
Project Suggested by Roman Bezrukavnikov\\ \; \\
July 31, 2019
}

 \begin{titlepage}
  \maketitle
  \begin{abstract}
    We consider the Iwahori-Hecke algebra of the symmetric group on $2n + r$ letters with parameter $q \in k^\times$.
    Let $e$ be the smallest integer such that $\brk{e}_q = 0$, or set $e = \infty$ if none exist.
    We modify Khovanov's Crossingless Matchings to include $2n$ nodes and $r$ anchors, and prove that the corresponding module is isomorphic to the Specht module $S^{(n+r,n)}$ when the specht module is irreducible and $e > n$.
    Additionally, we prove heuristics in support of the general case.
    Lastly, when $e = 5$, we prove an isomorphism between $D^{(n+r,n)}$ with $r \leq 3$ and a subrepresentations of Jordan-Shor's fibonacci representation.
  \end{abstract}

\tableofcontents

\end{titlepage}

@@

\fakesubsection{Crossingless Matchings}

\begin{definition}
  A \emph{crossingless matching on $2n+r$ indices with $r$ anchors} is a partition of $\cbr{1,\dots,2n+r}$ into $n$ parts of size $2$ and $r$ of size 1 such that no two parts of size two``cross'', i.e. there are no parts $(a,a')$ and $(b,b')$ such that $a < b < a' < b'$, and no parts of size one are ``inside'' of a part of size two, i.e. there are no $c, (a,a')$ such that $a < c < a'$.
  We will call these arcs and anchors, respectively.
  Then, define $W^r_{2n+r}$ to be the $k$-vector space with basis the set of generalized crossingless matchings on $2n+r$ indices with $r$ anchors.

@@

Note that the representations $W_{0 + r}^r$ and $S^{(r)}$ are isomorphic to the sign representation; 
we will prove that $W := W_{2n+r}^r$ and $S := S^{(n+r,n)'}$ are isomorphic as representations in the case that $e > n$ and $S^{(n+r,n)'}$ is irreducible.
Note that, when $r = 0$, these have the same dimension given by the $n$th catalan number $C_n$.

\def\vara{\alpha_1}
\def\varb{\alpha_2}
\def\varc{\varepsilon_1}
\def\vard{\delta}
\def\vare{\varepsilon_2}
\def\vs{\texttt{*}}
\def\vp{\texttt{0}}
\fakesubsection{Fibonacci Representation}
Now suppose that $e = 5$ and $k$ contains the algebraic number $(-1 - q^2 - q^3)^{3/2}$.
Let $V^m$ be a $k$-vector space with basis given by the strings $\cbr{\vs,\vp}^{n+1}$ such that the character $\vs$ never appears twice in a row. 
We will suppress the superscript whenever it is clear from context.

@@

% Acknowledgements should be given section-level heading in toc, subsection-level in text.
\addcontentsline{toc}{section}{\textbf{Acknowledgements}}
\fakesubsection{Acknowledgements}

\newpage

\section{Preliminaries on Specht Modules}
\subsection{Irreducibility of $S^\lambda$}
Let $k$ have characteristic $\ell$;
then, set
\[
  p := \begin{cases}
    \ell & \ell > 0\\
    \infty & \ell = 0
  \end{cases}.
\]  
For $h$ a natural number, let $\nu_p(h)$ be the $p$-adic evaluation of $h$.
As a convention, set $\nu_\infty(h) = 0$ for all $h$.
Define the function $\nu_{e,p} : \NN \rightarrow \cbr{-1} \cup \NN$ by
\[
  \nu_{e,p}(h) := \begin{cases}
    \nu_p(h) & e \mid h\\
    -1 & e \nmid h
  \end{cases}.
\]
Lastly, let $h_{ab}^\lambda$ be the hook length of node $(a,b)$ in $\brk \lambda$.

\begin{theorem}[James-Mathas]
  The following are equivalent:
  \begin{enumerate}
    \item $S^{\lambda} \simeq D^{\lambda}$.
    \item $\lambda$ is $e$-restricted and $S^\lambda$ is irreducible.
    \item $\nu_{e,p}\prn{h_{ab}^\lambda} = \nu_{e,p}\prn{h_{ac}^\lambda}$ for all nodes $(a,b)$ and $(a,c)$ in $\brk \lambda$.
  \end{enumerate}\qed
\end{theorem}

\begin{corollary}\label{S irreducibility}
  Assume $e > 2$ and $n > 0$.

  \begin{enumerate}[label={(\roman*)}]
    \item Suppose $e > n$.
      Then, $S^{(n+r,n)'}$ is irreducible iff $e \nmid l$ for all $r + 2 \leq l \leq n + r + 1$.
    \item Suppose $e \leq n$.
      Then, $S^{(n+r,n)'}$ is irreducible iff $e | r + 1$ and $\nu_{e,p}(h_{ab}^\lambda)$ acquires exactly two values over rows $a \leq n$.
  \end{enumerate}  
\end{corollary}
\begin{proof}
  \textbf{(i)}
  Note that $\nu_p(l) \neq -1$ for all $l$ and only hook lengths in the first column may vanish mod $e$;
  hence we may equivalently prove that $e$ divides no hook lengths in the first $n$ rows of the first column.
  These hook lengths are precisely $r + 2,\dots,n+r+1$.

@@

In the case that $\lambda = (n+r,n)$ and $e$ satisfy hypothesis \eqref{S irreducibility}.(i), say that $\lambda$ is \emph{$e$-top-indivisible}.

@@

\subsection{Branching}
\begin{theorem}[Kleshchev-Brundan]
  We have the following isomorphisms of vector spaces
  \begin{align*}
    \Homarg{\SH'}{S^\mu,\Res D^\lambda} 
    &\simeq \begin{cases}
      k & \mu \xrightarrow{\text{normal}} \lambda\\
      0 & \text{otherwise}
    \end{cases}\\
    \Homarg{\SH'}{D^\mu,\Res D^\lambda} 
    &\simeq \begin{cases}
      k & \mu \xrightarrow{\text{good}} \lambda\\
      0 & \text{otherwise}
    \end{cases}
  \end{align*}
  and $\Res D^\lambda$ is semisimple if and only if every normal number in $\lambda$ is good.\qed
\end{theorem}

\begin{corollary}
  Suppose $n,r > 0$.
  Then, we may characterize the socle of $\Res \, D^\lambda$ as follows:
  \[
    \soc\prn{\Res \, D^{(n+r,n)'}} \simeq \begin{cases}
      D^{(n + r - 1,n)'} & e \mid r+2\\
      D^{(n + r,n-1)'} & e \nmid r+2, \; e \mid r\\
      D^{(n+r-1,n)'} \oplus D^{(n+r,n-1)} & e \nmid r+2,r
    \end{cases}
  \]
\end{corollary}
\begin{proof}
  This amounts to computations of the hook lengths $\beta(1,2)$ and $\gamma(1,2)$:
  \begin{align*}
    \beta_\lambda(1,2) &= r + 2\\
    \gamma_\lambda(1,2) &= r
  \end{align*}
  Since $2$ is the largest removable number, $D^{(n+r,n-1)'} \subset D^{(n+r,n)'}$ iff $e \nmid r + 2$.
  Further, if $e \nmid r + 2$, then $D^{(n+r-1,n)'} \subset D^{(n+r,n)'}$ iff 1 is good iff $e \nmid r$
\end{proof}

@@

\section{Crossingless Matchings and Specht Modules}
\subsection{Irreducibility of $M$}
\begin{lemma}
  \label{Cyclic}
  Every basis vector in $W_{2n + r}^r$ is cyclic.
\end{lemma}
\begin{proof}
  We have already proven this in the $r = 0$ case, so suppose that $r > 0$.

@@

Let $K := \bigcap_{i=1}^{2n+r-1} \ker (1 + T_i) = \ker \bigoplus_{i = 1}^{2n + r - 1} (1 + T_i)$.
This will be a large technical tool in our proof of irreducibility.

\iffalse
\begin{lemma}\label{Represented}
  Let $w_j$ be the basis vector with anchors $1,\dots,r$ and all arcs of maximal length.
  Suppose $w \in K \backslash \cbr 0$.
  Then, $w_j$ is represented in $w$.
\end{lemma}
\begin{proof}
  We will show this in steps; first, we show that, given that a vector is represented with anchors $1,\dots,s$, there must be a vector represented in $w$ with $(s+1)$st anchor, including when $s = 0$;
  this implies that a vector is represented with anchors $1,\dots,r$.
  Then, we will show that, given a vector is represented with anchors $1,\dots,r$ and first $s$ arc-lengths $n,n-2,\dots,n-2s$, there is a vector represented with these and the $(s+1)$st arc-length $n-2s-2$.
  This implies that $w_j$ is represented.

@@

  \vspace{\lemsep}
  \textit{Step 2.}
  This step is similar;
  suppose that $s < n$ is the maximal number such that a vector with anchors $1,\dots,r$ and first $s$ arc-lengths $n,\dots,n-2s$ is represented.
  Take the vector $w_i$ which, among vectors represented in $w$ with anchors $1,\dots,r$ and first $s$ arc-lenths $n,\dots,n-2s$, has maximal length $t$ of the arc beginning at index $r + s + 1$.
  Then, $q^{1/2}(1 + T_{r + s + t})w_i$ is mapped to only by $w_i$ and vectors having anchors $1,\dots,r$ and first $s+1$ arc-lengths $n,\dots,n-2s,t'$ with $t' > t$, which are not represented in $w$;
  hence $q^{-1/2}(1 + T_{r + s + t})w_i$ is represented in $(1 + T_{r+s+t})w$, giving $w \notin \ker (1 + T_{r + s + t})$.
  The $s = 0$ case is similar, and implies that $s = n$.
\end{proof}
\fi

\begin{lemma}\label{Trivial}
  Suppose $e \nmid n + r + 1$.
  Then, $K = 0$.
\end{lemma}
\begin{proof}[Proof 1.]
  Consider the matrix $A = \bigoplus (1 + T_i)$ having kernel $K$.
  It is sufficient by lemma \ref{Represented} to show that $A$ includes a row $[0,\dots,0,1,0,\dots,0]$ with a nonzero entry only on the column $j$.

@@

  We may prove invertibility of this matrix by proving that $\det B_{n + r} = \brk{n + r + 1}_q$ inductively on $n+r$.
  This is satisfied for our base case $n + r = 1$, so suppose that it is true for each $m < n + r$.
  Then,
  \begin{align*}
    \det B_{n+r} &= (q + 1) \det B_{n + r - 1} - q \det B_{n + r - 2}\\
    &= (q + 1)(1 + \dots + q^{n + r - 1}) - (q + \dots + q^{n + r - 1})\\
    &= 1 + \dots + q^{n + r}\\
    &= \brk{n + r + 1}_q.
  \end{align*}
  Hence $\det B_{n + r} \neq 0$, and $K = 0$.
\end{proof}
\begin{proposition}
  The representation $W_{2n + r}^r$ is irreducible when $e > n + r + 1$. 
\end{proposition}
\begin{proof}
  We proceed by induction on $2n + r$.
  Note that, by identification with the trivial and sign representations, the base case $2n + r = 2$ is already prove, so suppose we have proven this for each $2m + s < 2n + r$.

@@

\subsection{Correspondence}
\begin{proposition}\;

  \begin{enumerate}[label={(\roman*)}]
    \item
    Suppose that $n,r > 0$.
    Then, a filtration of $\Res W_{2n + r}^r$ is given by
    \begin{equation}
      0 \subset W_{2n + r - 1}^{r-1} \subset \Res \, W_{2n + r}^r \label{Filtration}
     \end{equation}
     with $\Res \, W_{2n + r}^r / W_{2n + r - 1}^{r-1} \simeq W_{2n + r - 1}^{r + 1}$. 
    \item
      We have the following isomorphism of representations:
      \begin{equation}
        W_{2n - 1}^1 \simeq \Res \; W_{2n}^0 \label{0 Restriction}
       \end{equation}
  \end{enumerate}
    When the case is type (i) from the irreduibility lemma, this is a composition series.
\end{proposition}
\begin{proof}
  \textbf{(i)}
  Note that we may identify the subrepresentation of $\Res \; W_{2n + r}^r$ having anchor $n$ with $W_{2n + r - 1}^{r-1}$.

@@

\subsection{Kernels and Further Work}
\begin{lemma}\label{Represented}
  Let $w_j$ be the basis vector with anchors $1,\dots,r$ and all arcs of maximal length.
  Suppose $w \in K \backslash \cbr 0$.
  Then, $w_j$ is represented in $w$.
\end{lemma}
\begin{proposition}
  Suppose $e = n + r + 1$.
  Then $K$ is nontrivial.
\end{proposition}

@@

\bibliography{RepBib}
\iffalse
\begin{thebibliography}{9}

@@

\end{thebibliography}
\fi
